---
title: "Lab 05: Data Wrangling for Tables"
author: "Alison Hill & Steven Bedrick"
subtitle: "BMI 5/625"
always_allow_html: true
output:
  html_document:
    number_sections: yes
    theme: flatly
    toc: yes
    toc_depth: 2
    toc_float: yes
---

```{r,echo=FALSE,message=FALSE,warning=FALSE}
library(tidyverse)
library(dplyr)
library(janitor)
library(broom)
# devtools::install_github("ismayc/pnwflights14")
library(pnwflights14)
data("flights", package = "pnwflights14")
```


# Data-wrangling and tidying for tables

Making tables is often when we stretch our `dplyr` skills. Here are some additional more advanced examples that you may find helpful for today's assignment!

We'll use the `pnwflights14` package to practice our `dplyr` skills. We need to download the package from github using `devtools`.

```{r eval = FALSE}
# once per machine
install.packages("devtools")
devtools::install_github("ismayc/pnwflights14")
```

Now, we need to load the `flights` dataset from the `pnwflights14` package.

```{r eval = FALSE}
# once per work session
library(pnwflights14)
data("flights", package = "pnwflights14")
```

# More Fun `dplyr` Tricks

## `dplyr::select`

Recall that we use select to specify which columns in a dataframe you'd like to keep **by name**:

```{r use-dplyr-select}
# keep these 2 cols
mini_flights <- flights %>% 
  select(carrier, flight)
glimpse(mini_flights)

# keep first five cols
first_five <- flights %>% 
  select(year, month, day, dep_time, dep_delay)
glimpse(first_five)

# alternatively, specify range
first_five <- flights %>% 
  select(year:dep_delay)
glimpse(first_five)
```

We can also choose the columns we want by negation, that is, you can specify which columns to drop instead of keep. This way, all variables **not** listed are kept.

```{r drop-select}
# we can also use negation
all_but_year <- flights %>% 
  select(-year)
glimpse(all_but_year)
```



`dplyr::select` comes with several other helper functions...


```{r dplyr-select-helpers}
depart <- flights %>%
  select(starts_with("dep_"))
glimpse(depart)

times <- flights %>%
  select(contains("time"))
glimpse(times)

# note that we are not actually saving the new dataframe here
flights %>%
  select(-contains("time")) %>% head()


delays <- flights %>%
  select(ends_with("delay"))
glimpse(delays)
```

One of my favorite select helper functions is `everything()`, which allows you to use select to keep **all** your variables, but easily rearrange the columns without having to list all the variables to keep/drop.

```{r dplyr-select-everything}
new_order <- flights %>% 
  select(origin, dest, everything())
head(new_order)

# with negation
new_order2 <- flights %>% 
  select(origin, dest, everything(), -year)
head(new_order2)
```


We can also rename variables within select.

```{r rename-select}
flights2 <- flights %>%
  select(tail_num = tailnum, everything())
head(flights2)
```

If you don't want to move the renamed variables within your dataframe, you can use the `rename` function.

```{r rename}
flights3 <- flights %>%
  rename(tail_num = tailnum)
glimpse(flights3)
```

## `dplyr::filter`

As we have previously seen, `filter` has many flexible uses:


```{r dplyr-filter}
# flights taking off from PDX
pdx <- flights %>% 
  filter(origin == "PDX")
head(pdx)

# january flights from PDX
pdx_jan <- flights %>% 
  filter(origin == "PDX", month == 1) # the comma is an "and"
head(pdx_jan)

# flights to ATL (Atlanta) or BNA (Nashville)
to_south <- flights %>% 
  filter(dest == "ATL" | dest == "BNA") %>% # | is "or"
  select(origin, dest, everything())
head(to_south)

# flights from PDX to ATL (Atlanta) or BNA (Nashville)
pdx_to_south <- flights %>% 
  filter(origin == "PDX", dest == "ATL" | dest == "BNA") %>% # | is "or"
  select(origin, dest, everything())
head(pdx_to_south)

# alternatively, using group membership
south_dests <- c("ATL", "BNA")
pdx_to_south2 <- flights %>% 
  filter(origin == "PDX", dest %in% south_dests) %>% 
  select(origin, dest, everything())
head(pdx_to_south2)

# flights delayed by 1 hour or more
delay_1plus <- flights %>%
  filter(dep_delay >= 60)
head(delay_1plus)

# flights delayed by 1 hour, but not more than 2 hours
delay_1hr <- flights %>%
  filter(dep_delay >= 60, dep_delay < 120)
head(delay_1hr)
range(delay_1hr$dep_delay, na.rm = TRUE)

# even more efficient using between (always inclusive)
delay_bwn <- flights %>%
  filter(between(dep_delay, 60, 119))
head(delay_bwn)
range(delay_bwn$dep_delay, na.rm = TRUE)
```


## `dplyr::arrange`

```{r}
# default is ascending order
flights %>% 
  arrange(year, month, day) %>% head(n=20)

# descending order
flights %>% 
  arrange(desc(year), desc(month), desc(day)) %>% head(n=20)
```


## `dplyr::distinct`


```{r}
# unique origin-dest combinations
flights %>% 
  select(origin, dest) %>% 
  distinct %>% head(n=50)

# all unique destinations from PDX (there are 49, so we'll just list the first few)
from_pdx <- flights %>% 
  filter(origin == "PDX") %>% 
  select(origin, dest) %>%
  distinct(dest)
head(from_pdx)
```



## `dplyr::mutate`

`mutate` is used to transform existing variables or create new ones; here, we are using it to create an indicator variable that identifies flights that were entirely on-time.

```{r}
# add total delay variable
flights %>%
  mutate(tot_delay = dep_delay + arr_delay) %>%
  select(origin, dest, ends_with("delay"), everything()) %>% 
  head

# flights that were delayed at departure had on time or early arrivals?
arrivals <- flights %>%
  mutate(arr_ok = ifelse(dep_delay > 0 & arr_delay <= 0, 1, 0)) %>% 
  select(origin, dest, ends_with("delay"), carrier, arr_ok)

# peek at it
arrivals %>%
  filter(arr_ok == 1) %>%
  head
```



## `dplyr::summarise` (or `dplyr::summarize`)

We have seen `summarise` used to calculate summary statistics, like so:

```{r}
flights %>%
  summarise(mean(dep_delay, na.rm = TRUE))
```

But this can get tedious when we want to compute several summary statistics, like so:

```{r}
# we can also name that variable, and summarise multiple variables
flights %>%
  summarise(mean_delay = mean(dep_delay, na.rm = TRUE),
            sd_delay = sd(dep_delay, na.rm = TRUE),
            median_delay = median(dep_delay, na.rm = TRUE))
```

`summarise` together with `across()` lets us automate this process a bit:


```{r}
flights %>%
  filter(!is.na(dep_delay)) %>%
  select(dep_delay) %>%
  summarise(
    across(everything(), 
           lst(mean, sd, median)) # making this a list gets us nice column names
    )
  

# same thing over all numeric columns
flights %>%
  filter(!is.na(dep_delay)) %>%
  summarise(across(where(is.numeric), lst(mean, sd, median)))

# combine with pivot_longer, change names too
flights %>%
  filter(!is.na(dep_delay)) %>%
  summarise(across(dep_delay, lst(mean, stdev=sd, median))) %>% 
  pivot_longer(everything(), names_to="delay_stat", values_to="value", names_prefix = "dep_delay_")
```

We can use aggregating functions in `summarise`

```{r}
# how many unique destinations?
summary_table <- flights %>% 
  summarise(tot_flights = n(),
            tot_planes = n_distinct(tailnum),
            tot_carriers = n_distinct(carrier),
            tot_dests = n_distinct(dest),
            tot_origins = n_distinct(origin))

summary_table
```

The resulting table is a bit messy; we can reprocess using `pivot_longer` and `separate`:

```{r}
# chain with tidyr functions
summary_table %>% 
  pivot_longer(everything(), names_to="key", values_to="value") %>% 
  separate(key, into = c("tot", "entity")) %>% 
  select(-tot, total = value)
```
Or we can do it in a single call to `pivot_longer()`:

```{r}
summary_table %>% 
  pivot_longer(
    everything(), # grab all columns
    names_to="entity", # make a new column called "entity" to store the old column names
    names_prefix="tot_", # remove the "tot_" prefix from the old column names
    values_to="total") # put the values in a column called "total"
```


# `tidyr`

We'll work with a made up dataframe:

```{r}
df <- data.frame(
  id = 1:10,
  date = as.Date('2015-01-01') + 0:9,
  q1_m1_w1 = rnorm(10, 0, 1),
  q1_m1_w2 = rnorm(10, 0, 1),
  q1_m2_w3 = rnorm(10, 0, 1),
  q2_m1_w1 = rnorm(10, 0, 1),
  q2_m2_w1 = rnorm(10, 0, 1),
  q2_m2_w2 = rnorm(10, 0, 1)
)
```

```{r}
# HLO
head(df)
glimpse(df)
```

## `tidyr::pivot_longer`

First, let's pivot...
```{r}
df_tidy <- df %>%
  pivot_longer(q1_m1_w1:q2_m2_w2, names_to="key", values_to="value")
head(df_tidy)
```


## `tidyr::separate`

```{r}
# separate 1 col into 3 cols
df_sep <- df_tidy %>%
  separate(key, into = c("quarter", "month", "week"))
head(df_sep)

# separate 1 col into 2 cols
df_sep2 <- df_tidy %>%
  separate(key, into = c("quarter", "period"), extra = "merge")
head(df_sep2)
```
stringr vs. tidyr separate by regular expression

## `tidyr::extract`

`Extract` is essentially the same as `separate`, let's see how...

```{r}
# extract
df_ext <- df_sep2 %>%
  extract(period, into = "month")
head(df_ext)
```

We can see that, by default, it's taking the first alphanumeric chunk of the value, and `extract`ing it. We can specify much more complex patterns: 

```{r}
# this gives us same output as separate
df_ext <- df_sep2 %>%
  extract(period, into = c("month", "week"), 
          regex = "([[:alnum:]]+)_([[:alnum:]]+)")
head(df_ext)
```

## `tidyr::unite`

The `unite()` function lets us reverse the process:

```{r}
# let's say we want to combine quarter and month with an underscore
df_uni <- df_sep %>%
  unite(period, quarter:month) # sep = "_" is the default arg
head(df_uni)

# let's say we want to combine quarter and month with nothing
df_uni <- df_sep %>%
  unite(period, quarter:month, sep = "")
head(df_uni)
```

## `tidyr::pivot_wider`

```{r}
# finally let's spread
df_spread <- df_uni %>%
  pivot_wider(names_from=week, values_from=value)  # fill = NA is default arg
head(df_spread)
```

## Putting it all together (`pivot_longer() %>% separate() %>% pivot_wider()`)

All in one, if we had wanted to end up with one row per month:

```{r}
df_tidiest <- df %>% 
  pivot_longer(q1_m1_w1:q2_m2_w2, names_to="key", values_to="value") %>% 
  separate(key, into = c("quarter", "month", "week")) %>% 
  pivot_wider(names_from=week, values_from=value)
head(df_tidiest)
```

# `broom`

"The broom package takes the messy output of built-in functions in R, such as `lm`, `nls`, or `t.test`, and turns them into tidy data frames." So, broom tidies output from other R functions that are un-tidy.

See here for list of functions: https://github.com/dgrtwo/broom

Vignette: ftp://cran.r-project.org/pub/R/web/packages/broom/vignettes/broom.html

```{r}
fit <- lm(mpg ~ qsec + factor(am) + wt + factor(gear), 
          data = mtcars)
```

Un-tidy output from `lm`
```{r}
summary(fit)
```

Tidy output from `broom`
```{r}
tidy(fit)
```

This is very helfpul for making e.g. regression tables for papers!