---
title: "Bonus \"Lab 11\": JSON"
author: "Steven Bedrick"
date: "2025-06-04"
output: html_document
---

# Introduction

This walkthrough will illustrate a potential workflow for reproducing the data from the following paper:

> Axtell B, Munteanu C. [Tea, Earl Grey, Hot: Designing Speech Interactions from the Imagined Ideal of Star Trek.](https://dl.acm.org/doi/abs/10.1145/3411764.3445640) In: Proceedings of the 2021 CHI Conference on Human Factors in Computing Systems. Yokohama Japan: ACM; 2021. p. 1–14.

The associated data can be found [here](https://www.speechinteraction.org/TNG/) (and is on Posit.cloud).

# Setup

```{r,message=FALSE,error=FALSE}
library(tidyverse)
library(here)
library(jsonlite)
library(gt)
```

# Overall workflow:

1. Know your data
2. Think about desired final structure
3. Read JSON into gnarly list
4. Work from the outside inwards.

# Know your data:

Check in text editor (or JSON-viewer of your choice- Firefox will work as well)


```json
{
	"102": {
		"255": {
			"char": "Tasha",
			"line": "Battle bridge.",
			"direction": "The doors snap closed and the lift moves. Riker looks Tasha over  waiting  then:",
			"type": [
				"Statement"
			],
			"pri_type": "Statement",
			"domain": [
				"IoT"
			],
			"sub-domain": [
				"Turbolift"
			],
			"nv_resp": true,
			"interaction": "Battle bridge.",
			"char_type": "Person",
			"is_fed": true,
			"error": false
		},
		...
```

Outer-most part of the document is a dict; keys are episodes, values are dictionaries containing utterance IDs as keys and a dictionary as a value. Each of _these_ has the actual information about the utterance.

# Thinking about desired final structure:

Since we want a final tally of interaction type by speaker type, we'll want something like this:

Each row should be an utterance, with columns for:

1. Episode ID
2. Utterance ID within that episode
3. The category of speaker (person or computer)
4. The type of interaction (command, wake word, etc.)

Looking at the structure of our data, it looks like the important inner keys are going to be `type` and `char_type`.

# Initial loading

`jsonlite::read_json()` is a good place to start; note that for _some_ JSON files, it can automatically do a fair bit of the work that we are about to do by hand. If your JSON file is relatively simple, and does not have deeply-nested objects with varying dimensions, `as_tibble()` may be able to coerce the nested list into a dataframe... but not always, and not for _this_ JSON file, because of the amount of heterogeneity in the data. 
So we will be doing this "by hand".

```{r}
j_path <- here("data/lab11/teaearlgreyhotdataset.json")
j <- read_json(j_path)
```

Note what we get back from `read_json`: 

```{r}
typeof(j)
length(j)
```

Why do we get a list and not a data frame? Because JSON allows for _arbitrarily complicated_ nested structures, and so we don't have any guarantee that what's in that file will be amenable to flattening without some very file-specific work. So `jsonlite` punts on the issue and makes a list-of-lists for us, so we can deal with it ourselves.

Before we continue, note that this file is pretty big. To make it a little easier to work with while we're just getting started, I am going to make a small version that only has two episodes' worth of data. I'll do this "by hand" outside of R, in my text editor.

```{r}
j_path <- here("data/lab11/teaearlgreyhotdataset.small.json")
j <- read_json(j_path)
length(j)
```

This is a _named_ list, so we can access its elements by name:

```{r}
names(j)
```


```{r}
length(j$"102")
```

And the sub-lists are themselves named:

```{r}
names(j$"102")
```

Look back at the JSON file itself; you'll see that these correspond to the dictionary keys in the file.

It's turtles all the way down, to the final layer of the information objects themselves:

```{r}
names(j$"102"$"255")
```

```{r}
j$"102"$"255"$"char"
```


# Work from the outside in

Now, we are going to work our way through the document, flattening and filtering as we go, until we end up with a tidy data frame. 
Step one: turn our list into a very simple DF, using `enframe()`:

```{r}
j2 <- j %>% enframe
j2
```

See what happened here? We started with a named list, and ended up with a dataframe where one column is the names of the list and the second column is the values. 

Next, we're going to go one more level down, but before we do, let's rename our columns- by default, `enframe` just gives the uninformative column names "name" and "value", which (as you will see) can quickly become confusing We can do this using `rename()`, or we can override those defaults when we call `enframe`. 

```{r}
j2 <- j2 %>% rename(episode.id=name, episode=value) 
j2
```

Now that we've got useful names, let's go one level deeper (Inception-style) and turn our episodes into nested dataframes:

```{r}
j3 <- j2 %>% mutate(episode=purrr::map(episode,enframe))
j3
```

Now, if we `unnest` the episode column...

```{r}
j3 <- j3 %>% unnest(episode)
j3
```

Now, see what we've got? A tidy data frame, one row per utterance, with the default column names from `enframe`. Before we forget, let's rename our columns:

```{r}
j3 <- j3 %>% rename(utterance.id=name, utterance=value)
j3
```

I bet you can guess what comes next: that's right, we are going to `enframe` our utterances, and unnest!

```{r}
j4 <- j3 %>% mutate(utterance=purrr::map(utterance,enframe))
j4
```

When we unnest _this_ time, let's see what happens:

```{r}
j5 <- j4 %>% unnest(utterance)
j5
```

Woah! Now we've got one line for each _key_ in the inner-most part of the JSON file, with a corresponding list value. A few things to note:

1. The `value` column here is of type `list`, but note that the different rows have different types of lists in them - some have characters, some have logicals, some have lists, etc.
2. Some of those lists have different numbers of entries- most are 1, but some are 0. 
3. The actual data we want (the speaker, the interaction type, etc.) is still locked away in a list.

Our next step will be to pivot this into a slightly wider dataframe:

```{r}
j6 <- j5 %>% pivot_wider(names_from=name, values_from=value)
j6
```

This is starting to look like we're getting closer to a dataframe we can work with! 
We still have a little bit of work to do, however.
First, let's make our lives simpler by getting rid of some of the columns we don't need:

```{r}
j6 <- j6 %>% select(episode.id, utterance.id, char_type, type)
j6
```

Note the length of some of the entries in the `type` column- utterance 363 in espisode 102 has length 3. Looking at the JSON file, we can see that this is because this utterance was assigned three types- "Wake word", "Conversation", and "Question".

Because the "type" column is itself a container JSON element (a list), we actually will need to `unnest` it twice. This is the sort of thing that `jsonlite` has helpers to do, but for a simple file format like this one I personally usually just do it all by hand, so I can be sure about what's going on.

```{r}
j7 <- j6 %>% unnest(char_type) %>% unnest(type) %>% unnest(type)
j7
```
At this point, we've got a nice, tidy data frame, and we are ready to use regular methods to compute our table:

```{r}
j7 %>% group_by(char_type, type) %>% 
  summarise(n=n()) %>% 
  mutate(denom=sum(n)) %>% 
  mutate(prop=n/denom) %>% 
  select(char_type,  type, prop) %>% 
  pivot_wider(names_from=char_type, values_from=prop) %>% 
  select(type, Person, Computer) %>% 
  arrange(-Person) 
```

And we're all set! Now we just need to repeat the above workflow, pointed at the true file:

```{r, echo=FALSE}
full_json_path <- here("data/lab11/teaearlgreyhotdataset.json")
full_j <- read_json(full_json_path)
full_j %>% enframe %>% 
  rename(episode.id=name, episode=value) %>% 
  mutate(episode=purrr::map(episode,enframe)) %>% 
  unnest(episode) %>% 
  rename(utterance.id=name, utterance=value) %>% 
  mutate(utterance=purrr::map(utterance,enframe)) %>% 
  unnest(utterance) %>% 
  pivot_wider(names_from=name, values_from=value) %>% 
  select(episode.id, utterance.id, char_type, type) %>% 
  unnest(char_type) %>% unnest(type) %>% unnest(type) %>% 
   group_by(char_type, type) %>% 
  summarise(n=n()) %>% 
  mutate(denom=sum(n)) %>% 
  mutate(prop=n/denom) %>% 
  select(char_type,  type, prop) %>% 
  pivot_wider(names_from=char_type, values_from=prop, values_fill=0.0) %>% 
  select(type, Person, Computer) %>% 
  arrange(-Person, -Computer) %>% 
  gt %>% fmt_percent(columns=c(Person, Computer))
  
```

Note that these numbers don't match what is in the paper; I suspect that what we could be seeing is that the authors may have used the `pri_type` field instead of the `type` field. Repeating our analysis, looking for that field, we see:

```{r}
full_j %>% enframe %>% 
  rename(episode.id=name, episode=value) %>% 
  mutate(episode=purrr::map(episode,enframe)) %>% 
  unnest(episode) %>% 
  rename(utterance.id=name, utterance=value) %>% 
  mutate(utterance=purrr::map(utterance,enframe)) %>% 
  unnest(utterance) %>% 
  pivot_wider(names_from=name, values_from=value) %>% 
  select(episode.id, utterance.id, char_type, pri_type) %>% 
  unnest(c(char_type, pri_type)) %>% 
   group_by(char_type, pri_type) %>% 
  summarise(n=n()) %>% 
  mutate(denom=sum(n)) %>% 
  mutate(prop=n/denom) %>% 
  select(char_type,  pri_type, prop) %>% 
  pivot_wider(names_from=char_type, values_from=prop, values_fill=0.0) %>% 
  select(pri_type, Person, Computer) %>% 
  arrange(-Person, -Computer) %>% 
  gt %>% fmt_percent(columns=c(Person, Computer))

```


This _still_ doesn't match what is in the paper, and I think it's because I was mistaken about the denominator. 

```{r}
full_j %>% enframe %>% 
  rename(episode.id=name, episode=value) %>% 
  mutate(episode=purrr::map(episode,enframe)) %>% 
  unnest(episode) %>% 
  rename(utterance.id=name, utterance=value) %>% 
  mutate(utterance=purrr::map(utterance,enframe)) %>% 
  unnest(utterance) %>% 
  pivot_wider(names_from=name, values_from=value) %>% 
  select(episode.id, utterance.id, char_type, type) %>%
  unnest(c(char_type, type)) %>% unnest(type)  %>%

  janitor::tabyl(type, char_type) %>% data.frame %>% 
  mutate(Comp.prop=Computer/sum(Computer), Person.prop=Person/sum(Person)) %>% arrange(-Person.prop, -Comp.prop) %>% 
  select(type,Person=Person.prop, Computer=Comp.prop) %>% 
  gt
```

This _still_ isn't in alignment with what the paper reported, but at this point you get the idea. :-D



